The Units of Permissionless Consensus: \\ Towards Mobile and Edge Computing

### 2022-11-09 theoretical-consensus/article/article.tex
clean up tex

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. The IEEE's font choices and paper sizes can   ***
% *** trigger bugs that do not appear when using other class files.       ***                          ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

@@

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Line breaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{The Units of Permissionless Consensus: \\ Towards Mobile and Edge Computing}
%
%
% author names and IEEE memberships
% note positions of commas and non breaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

@@

% note the % following the last \IEEEmembership and also \thanks - 
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
% 
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!
%
% a space would be appended to the last name and could cause every name on that
% line to be shifted left slightly. This is one of those "LaTeX things". For
% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
% "AB" then you have to do: "\textbf{A}\textbf{B}"
% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names. For what it is worth,
% this is a minor point as most people would not even notice if the said evil
% space somehow managed to creep in.

% The paper headers
% \markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
% {Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for IEEE Journals}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

% If you want to put a publisher's ID mark on the page you can do it like
% this:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2015 IEEE}
% Remember, if you use this you must call \IEEEpubidadjcol in the second
% column for its text to clear the IEEEpubid mark.

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
Disclaimer: This paper is a work in progress. It is not a final version and it is not intended to be submitted yet to the Distributed Systems Seminar. The content of this paper is subject to change.
\end{abstract}

% Note that keywords are not normally used for peer review papers.
\begin{IEEEkeywords}
IEEE, IEEEtran, journal, \LaTeX, paper, template.
\end{IEEEkeywords}

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
% \IEEEpeerreviewmaketitle

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by the IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{L}{ong} has been the time when consensus started 
to be defined as a fundamental problem of distributed systems \cite{pease1980reaching, lamport2019byzantine, lamport1983weak}. 
Generally, consensus means reaching an agreement between multiple parties 
in the potential presence of faulty individuals. As per multi-agent systems, 
interacting over computer networks, consensus is thought to be the result 
of a coordination effort, such that those parties agree on some value at a 
given moment. Achieving consensus implies that the system shall be reliable 
and fault-tolerant. However, the consensus problem has been limited by some 
assumptions on the networks. The well-known "secure Byzantine-Fault-Tolerant 
multiparty consensus systems" that have been designed over the years are usually 
meant to work only with a set of known participants, faulty or not \cite{castro1999practical}. The 
other side of the coin is the permissionless consensus challenge, consisting of 
achieving agreement in an environment where the participants are unknown and 
untrusted \cite{nakamoto2008bitcoin, buterin2014next}. Plus, there are other intrinsic particularities of 
this type of networks, for example, their openness, and the lack of any kind of central 
authority. This adds another layer of complexity to the problem, as the participants 
are not only unknown and untrusted but can also join or leave the network at any time, 
freely choosing if they want to participate in the consensus protocol or not. Nevertheless, 
the problem of permissionless consensus can still be seen as a special case of the more 
general consensus and can still be formalized in the same way. In this paper, we will 
focus on consensus in permissionless systems, especially in the context of blockchain 
networks. We will reason about its meaningfulness, ultimately by trying to identify the units that 
may underpin consensus. We will also discuss the current state-of-the-art, with a 
particular interest in the shift of the consensus layer of these distributed networks 
to mobile and edge computing environments, for which computationally expensive consensus 
algorithms are impractical and unfair.

@@

There have been several attempts to address the challenges of resource-constrained networks,
and to understand the limitations of the existing consensus protocols in this context.
A common conclusion is that the existing protocols are not suitable for these networks,
due either to the high resource consumption, or to the centralization of various aspects of the protocol,
or also to high communication overheads, that are translated to high latency and low throughput \cite{SALIMITARI2020100212},
making these protocols unsuitable for resource-constrained applications. Other concerns are the
security issues, as some attempts to overcome the resource constraints may lead to new vulnerabilities,
which may be exploited by malicious and more powerful entities.

A thoroughly mentioned approach is the IOTA Tangle protocol \cite{popov2018tangle}, which
presents a friendly design for resource-constrained networks, infinitely scalable, and with low latency.
However, there are still known security and centralization limitations, as presented in the sections before.
Another concept that has been explored is sharding \cite{SALIMITARI2020100212, queralta2021blockchain}, which is a
technique that allows for the partitioning of the network into smaller sub-networks,
that are then managed by different entities, validating transactions in parallel. 
This approach may be used to overcome the scalability bottleneck,
but it may also introduce new issues that are still to be researched.
Frankly, there are indeed some well established protocols that, all in all, can be adapted to networks with
resource-constrained devices that are not necessarily short on the resources needed to run the protocols.
For example, a lot of \emph{PoS} related protocols may be adapted to resource-constrained networks, where 
the stake is the only considerable resource that the nodes need. There are a multitude of projects that are
currently exploring this approach, for example, with the Ethereum transition to \emph{PoS} \cite{buterin2017casper}.
However, this does not address the hipothesis of taking advantage of the very nature of these limited environments, 
as simply choosing a BFT protocol that is resource-efficient
may also put us far apart from the ultimate decentralization and scalability goal of 
other and fairer means of consensus participation.
An interesting idea is based on the use of \emph{Proof of Location} for 
coordinating the consensus. The easiness of movement and targetted location deployment
of a lot of these low-power devices may be used to achieve a decentralized consensus
protocol, where elected nodes may be somehow chosen based on their location, 
either by proximity, or by some other withnessing mechanism. Variations and 
extensions of this idea, with different and wide use cases, have been seen in
\cite{foam2018location, xyo2018location}

@@

% Subsection text here.

% needed in second column of first page if using \IEEEpubid
%\IEEEpubidadjcol

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation results for the network.}
%\label{fig_sim}
%\end{figure}

% Note that the IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command,
% and the \label for the overall figure must come after \caption.
% \hfil is used as a separator to get equal spacing.
% Watch out that the combined width of all the subfigures on a 
% line do not exceed the text width or a line break will occur.
%
%\begin{figure*}[!t]
%\centering
%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
%\label{fig_second_case}}
%\caption{Simulation results for the network.}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat[]), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.
% Be aware that for subfig.sty to generate the (a), (b), etc., subfigure
% labels, the optional argument to \subfloat must be present. If a
% subcaption is not desired, just leave its contents blank,
% e.g., \subfloat[].

% An example of a floating table. Note that, for IEEE style tables, the
% \caption command should come BEFORE the table and, given that table
% captions serve much like titles, are usually capitalized except for words
% such as a, an, and, as, at, but, by, for, in, nor, of, on, or, the, to
% and up, which are usually not capitalized unless they are the first or
% last word of the caption. Table text will default to \footnotesize as
% the IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that the IEEE does not put floats in the very first column
% - or typically anywhere on the first page for that matter. Also,
% in-text middle ("here") positioning is typically not used, but it
% is allowed and encouraged for Computer Society conferences (but
% not Computer Society journals). Most IEEE journals/conferences use
% top floats exclusively. 
% Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the
% \fnbelowfloat command of the stfloats package.

% \section{Conclusion}
% The conclusion goes here.

% if have a single appendix:
%\appendix[Proof of the Zonklar Equations]
% or
%\appendix  % for no appendix heading
% do not use \section anymore after \appendix, only \section*
% is possibly needed

% use appendices with more than one appendix
% then use \section to start each appendix
% you must declare a \section before using any
% \subsection or using \label (\appendices by itself
% starts a section numbered zero.)
%

% \appendices
% \section{Proof of the First Zonklar Equation}
% Appendix one text goes here.

% % you can choose not to have a title for an appendix
% % if you want by leaving the argument blank
% \section{}
% Appendix two text goes here.

% % use section* for acknowledgment
% \section*{Acknowledgment}

% The authors would like to thank...

% Can use something like this to put references on a page
% by themselves when using endfloat and the captionsoff option.
\ifCLASSOPTIONcaptionsoff
  \newpage
\fi

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/

%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)

@@

% biography section
% 
% If you have an EPS/PDF photo (graphicx package needed) extra braces are
% needed around the contents of the optional argument to biography to prevent
% the LaTeX parser from getting confused when it sees the complicated
% \includegraphics command within an optional argument. (You could create
% your own custom macro containing the \includegraphics command to make things
% simpler here.)
%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
% or if you just want to reserve a space for a photo:

% \begin{IEEEbiography}{Michael Shell}
% Biography text here.
% \end{IEEEbiography}

% % if you will not have a photo at all:
% \begin{IEEEbiographynophoto}{John Doe}
% Biography text here.
% \end{IEEEbiographynophoto}

% insert where needed to balance the two columns on the last page with
% biographies
%\newpage

% \begin{IEEEbiographynophoto}{Jane Doe}
% Biography text here.
% \end{IEEEbiographynophoto}

% You can push biographies down or up by placing
% a \vfill before or after them. The appropriate
% use of \vfill depends on what kind of text is
% on the last page and whether or not the columns
% are being equalized.

%\vfill

% Can be used to pull up biographies so that the bottom of the last one
% is flush with the other column.
%\enlargethispage{-5in}

>>>>

@@

\begin{document}

\title{The Units of Permissionless Consensus: \\ Towards Mobile and Edge Computing}

@@

\maketitle

\begin{abstract}
Disclaimer: This paper is a work in progress. It is not a final version and it is not intended to be submitted yet to the Distributed Systems Seminar. The content of this paper is subject to change.
\end{abstract}

\begin{IEEEkeywords}
IEEE, IEEEtran, journal, \LaTeX, paper, template.
\end{IEEEkeywords}

\section{Introduction}

\IEEEPARstart{L}{ong} has been the time when consensus started 
to be defined as a fundamental problem of distributed systems \cite{pease1980reaching, lamport2019byzantine, lamport1983weak}. 
Generally, consensus means reaching an agreement between multiple parties 
in the potential presence of faulty individuals. As per multi-agent systems, 
interacting over computer networks, consensus is thought to be the result 
of a coordination effort, such that those parties agree on some value at a 
given moment. Achieving consensus implies that the system shall be reliable 
and fault-tolerant. However, the consensus problem has been limited by some 
assumptions on the networks. The well-known "secure Byzantine-Fault-Tolerant 
multiparty consensus systems" that have been designed over the years are usually 
meant to work only with a set of known participants, faulty or not \cite{castro1999practical}. The 
other side of the coin is the permissionless consensus challenge, consisting of 
achieving agreement in an environment where the participants are unknown and 
untrusted \cite{nakamoto2008bitcoin, buterin2014next}. Plus, there are other intrinsic particularities of 
this type of networks, for example, their openness, and the lack of any kind of central 
authority. This adds another layer of complexity to the problem, as the participants 
are not only unknown and untrusted but can also join or leave the network at any time, 
freely choosing if they want to participate in the consensus protocol or not. Nevertheless, 
the problem of permissionless consensus can still be seen as a special case of the more 
general consensus and can still be formalized in the same way. In this paper, we will 
focus on consensus in permissionless systems, especially in the context of blockchain 
networks. We will reason about its meaningfulness, ultimately by trying to identify the units that 
may underpin consensus. We will also discuss the current state-of-the-art, with a 
particular interest in the shift of the consensus layer of these distributed networks 
to mobile and edge computing environments, for which computationally expensive consensus 
algorithms are impractical and unfair.

@@

There have been several attempts to address the challenges of resource-constrained networks,
and to understand the limitations of the existing consensus protocols in this context.
A common conclusion is that the existing protocols are not suitable for these networks,
due either to the high resource consumption, or to the centralization of various aspects of the protocol,
or also to high communication overheads, that are translated to high latency and low throughput \cite{SALIMITARI2020100212},
making these protocols unsuitable for resource-constrained applications. Other concerns are
the high space consumption, to store the older blockchain information, or the security issues, 
as some attempts to overcome the resource constraints may lead to new vulnerabilities,
which may be exploited by malicious and more powerful entities.

A thoroughly mentioned approach is the IOTA Tangle protocol \cite{popov2018tangle}, which
presents a friendly design for resource-constrained networks, infinitely scalable, and with low latency.
However, there are still known security and centralization limitations, as presented in the sections before \cite{8168250}.
Another concept that has been explored is sharding \cite{SALIMITARI2020100212, queralta2021blockchain}, which is a
technique that allows for the partitioning of the network into smaller sub-networks,
that are then managed by different entities, validating transactions in parallel. 
This approach may be used to overcome the scalability bottleneck,
but it may also introduce new issues that are still to be researched.
Frankly, there are indeed some well established protocols that, all in all, can be adapted to networks with
resource-constrained devices that are not necessarily short on the resources needed to run the protocols.
For example, a lot of \emph{PoS} related protocols may be adapted to resource-constrained networks, where 
the stake is the only considerable resource that the nodes need. There are a multitude of projects that are
currently exploring this approach, for example, with the Ethereum transition to \emph{PoS} \cite{buterin2017casper}.
However, this does not address the hypothesis of taking advantage of the very nature of these limited environments, 
as simply choosing a BFT protocol that is resource-efficient
may also put us far apart from the ultimate decentralization and scalability goal of 
other and fairer means of consensus participation.
An interesting idea is based on the use of \emph{Proof of Location} for 
coordinating the consensus. The easiness of movement and targetted location deployment
of a lot of these low-power devices may be used to achieve a decentralized consensus
protocol, where elected nodes may be somehow chosen based on their location, 
either by proximity, or by some other witnessing mechanism. Variations and 
extensions of this idea, with different and wide use cases, have been seen in
\cite{foam2018location, xyo2018location}

@@

\ifCLASSOPTIONcaptionsoff
  \newpage
\fi

@@
